---
title: "Pacific Herring spawn index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pacific Herring spawn index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette provides an example on how to calculate the spawn index for Pacific Herring (*Clupea pallasii*) in British Columbia, Canada.
Load the load the `SpawnIndex` package, and some additional pacakges used in this vignette.

```{r libraries, message = FALSE, warning = FALSE}
library(SpawnIndex)
library(tidyverse)
library(kableExtra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
# Black and white theme
theme_set(theme_bw())
# Canada: coarse polygons for mapping
canada <- ne_countries(scale = "large", returnclass = "sf", country = "canada")
```

# Set up

Specify the region and years of interest.
Also, load the parameter values required for spawn index calculations.

```{r pars}
# Parameters
region <- "WCVI" # Region: HG, PRD, CC, SoG, WCVI, A27, or A2W
yrRange <- 2010:2015 # Years (>= 1951)
data(pars) # Parameters for spawn index calculations
```

This example will focus on `r region` from `r min(yrRange)` to `r max(yrRange)`.

# Spatial information

Load the spatial information for this region;
we will use the example spawn database that comes with the package (`HerringSpawn.mdb`).
Note that this example database only includes data for WCVI from 2010 to 2015.

```{r areas}
# Database location
dbLoc <- system.file("extdata", package = "SpawnIndex")
# List indicating the necessary tables: sections and locations
areaLoc <- list(
  loc = dbLoc, db = "HerringSpawn.mdb",
  fns = list(sections = "Sections", locations = "Location")
)
# Load area data and remove areas with no spatial information
areas <- LoadAreaData(reg = region, where = areaLoc)
# Determine spatial extent (for maps)
xLims <- range(areas$Longitude)
yLims <- range(areas$Latitude)
```

In this example database, `r region` has
`r n_distinct(areas$StatArea)` Statistical Areas,
`r n_distinct(areas$Section)` Sections, and
`r n_distinct(areas$LocationName)` Locations.
Here we show the first 10 Locations.

```{r areasTable, echo = FALSE, results = 'asis'}
areas %>%
  select(Region, StatArea, Section, LocationName, Longitude, Latitude) %>%
  arrange(StatArea, Section, LocationName) %>%
  slice(1:10) %>%
  kable(
    caption = paste("Example spawn survey Locations in the", region, "region."),
    booktabs = TRUE
  )
```

# Spawn index calculations

Calculate the conversion factor to convert the number of eggs to the spawn index (i.e., biomass in tonnes).

```{r theta}
theta <- CalcEggConversion()
theta
```

Thus, we divide the number of eggs by `r theta` to estimate the biomass of fish that spawned.
This conversion factor is the default in the spawn index calculations below
(e.g., `CalcSurfSpawn()`).
Calculate the spawn index for the three spawn survey types:
surface, Macrocystis, and understory.

## Surface spawn index

Calculate the surface spawn index.
Surface spawn index calculations require a table to convert intensity categories to the number of egg layers.

```{r surf}
# Data
data(intensity) # Spawn intensity categories and number of egg layers
# List indicating the necessary tables
surfLoc <- list(
  loc = dbLoc, db = "HerringSpawn.mdb",
  fns = list(
    regionStd = "RegionStd", sectionStd = "SectionStd", poolStd = "PoolStd",
    surface = "tSSSurface", allSpawn = "tSSAllspawn"
  )
)
# Calculate the surface spawn index
surfSpawn <- CalcSurfSpawn(where = surfLoc, a = areas, yrs = yrRange)
```

Show surface spawn on a map, aggregated by Location and year.

```{r surfLocYr, fig.width=7, fig.height=7}
# Make a spatial object for mapping
surfLocYr <- surfSpawn$SI %>%
  group_by(Year, LocationCode) %>%
  summarise(Surface = round(sum(SurfSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = surfLocYr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Surface), size = 3) +
  scale_colour_viridis_c() +
  labs(title = "Surface", colour = "Spawn\nindex (t)") +
  coord_sf(xlim = xLims, ylim = yLims) +
  facet_wrap(Year ~ ., ncol = 2)
```

Calculate the total surface spawn by year.

```{r surfYr, results = 'asis'}
# Aggregate the surface spawn index by year
surfYr <- surfSpawn$SI %>%
  group_by(Year) %>%
  summarise(Surface = sum(SurfSI, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(Year)
# Table
kable(
  x = surfYr, digits = 1, caption = "Surface spawn index in tonnes (t) by year.",
  booktabs = TRUE
)
```

## Macrocystis spawn index

Calculate the Macrocystis spawn index.

```{r macro}
# List indicating the necessary tables
macroLoc <- list(
  loc = file.path(dbLoc), db = "HerringSpawn.mdb",
  fns = list(
    allSpawn = "tSSAllspawn", plants = "tSSMacPlant",
    transects = "tSSMacTrans"
  )
)
# Calculate the Macrocystis spawn index
macroSpawn <- CalcMacroSpawn(where = macroLoc, a = areas, yrs = yrRange)
```

Show Macrocystis spawn on a map, aggregated by Location and year.

```{r macroLocYr, fig.width=7, fig.height=7}
# Make a spatial object for mapping
macroLocYr <- macroSpawn$SI %>%
  group_by(Year, LocationCode) %>%
  summarise(Macrocystis = round(sum(MacroSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = macroLocYr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Macrocystis), size = 3) +
  scale_colour_viridis_c() +
  labs(title = "Macrocystis", colour = "Spawn\nindex (t)") +
  coord_sf(xlim = xLims, ylim = yLims) +
  facet_wrap(Year ~ ., ncol = 2)
```

Calculate the total Macrocystis spawn by year.

```{r macroYr, results = 'asis'}
# Aggregate the Macrocystis spawn index by year
macroYr <- macroSpawn$SI %>%
  group_by(Year) %>%
  summarise(Macrocystis = sum(MacroSI, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(Year)
# Table
kable(
  x = macroYr, digits = 1,
  caption = "Macrocystis spawn index in tonnes (t) by year.", booktabs = TRUE
)
```

## Understory spawn index

Calculate the understory spawn index.
Understory spawn index calculations require tables to correct spawn widths,
and to specify algae coefficients.

```{r under}
# Data
data(underWidthFac) # Understory spawn width correction factors
data(algaeCoefs) # Algae types and coefficients
# List indicating the necessary tables
underLoc <- list(
  loc = file.path(dbLoc), db = "HerringSpawn.mdb",
  fns = list(
    allSpawn = "tSSAllspawn", algTrans = "tSSVegTrans",
    stations = "tSSStations", algae = "tSSVegetation"
  )
)
# Calculate the understory spawn index
underSpawn <- CalcUnderSpawn(where = underLoc, a = areas, yrs = yrRange)
```

Show understory spawn on a map, aggregated by Location and year.

```{r underLocYr, fig.width=7, fig.height=7}
# Make a spatial object for mapping
underLocYr <- underSpawn$SI %>%
  group_by(Year, LocationCode) %>%
  summarise(Understory = round(sum(UnderSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = underLocYr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Understory), size = 3) +
  scale_colour_viridis_c() +
  labs(title = "Understory", colour = "Spawn\nindex (t)") +
  coord_sf(xlim = xLims, ylim = yLims) +
  facet_wrap(Year ~ ., ncol = 2)
```

Calculate the total understory spawn by year.

```{r underYr, results = 'asis'}
# Aggregate the understory spawn index by year
underYr <- underSpawn$SI %>%
  group_by(Year) %>%
  summarise(Understory = sum(UnderSI, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(Year)
# Table
kable(
  x = underYr, digits = 1,
  caption = "Understory spawn index in tonnes (t) by year.", booktabs = TRUE
)
```

# Total spawn index

Calculate the total spawn index by year as the sum of the spawn index in the three survey types:
surface, Macrocystis, and understory.

```{r totalYr, results = 'asis'}
# Total spawn index
totalYr <- full_join(x = surfYr, y = macroYr, by = "Year") %>%
  full_join(y = underYr, by = "Year") %>%
  replace_na(replace = list(Surface = 0, Macrocystis = 0, Understory = 0)) %>%
  mutate(Total = Surface + Macrocystis + Understory) %>%
  arrange(Year)
# Table
kable(
  x = totalYr, digits = 1,
  caption = "Total spawn index in tonnes (t) by year.", booktabs = TRUE
) %>%
  add_header_above(header = c("", "Survey" = 3, ""))
```
