---
title: "Introduction to Pacific Herring spawn index calculations"
author: "Matthew H. Grinnell"
date: "`r Sys.Date()`"
output:
      rmarkdown::html_vignette:
        fig_caption: yes
        toc: true
        number_sections: true
bibliography:
  - "../inst/REFERENCES.bib"
  - "../inst/PACKAGES.bib"
csl: "../inst/cjfas.csl"
link-citations: true
vignette: >
  %\VignetteIndexEntry{Introduction to Pacific Herring spawn index calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5.75
)
options(knitr.kable.NA = "")
```

```{r citation, include = FALSE}
knitr::write_bib(x = "SpawnIndex", file = "../inst/PACKAGES.bib")
```

This vignette shows how to calculate the spawn index for
Pacific Herring (*Clupea pallasii*) in British Columbia, Canada.
See the technical report for background information [@GrinnellEtalYYYY].
Source code is on the
[GitHub repository](https://github.com/grinnellm/SpawnIndex).
Load the `SpawnIndex` package [@R-SpawnIndex], and
a few other packages for data wrangling, as well as making tables, and figures.

```{r libraries, message = FALSE, warning = FALSE}
# Packages
library(SpawnIndex)
library(dplyr)
library(ggplot2)
library(tidyr)
library(kableExtra)
library(sf)
library(rnaturalearth)
library(rnaturalearthhires)
# Suppress summarise info
options(dplyr.summarise.inform = FALSE)
```

We provide three examples:

1. Spawn index calculations using the example database,
1. Surface spawn index calculations using custom data, and
1. Spawn index calculations for spawn-on-kelp (SOK) operations.

# Spawn index calculations using the example database

Calculate the Pacific Herring spawn index using the example database
provided with the SpawnIndex package.

## Set up

Specify the region and years of interest.
Then load the parameters required for spawn index calculations, and
load some land polygons for maps.

```{r pars}
# Region
region <- "WCVI"
# Years
yr_range <- 2010:2013
# Set a random seed
set.seed(12)
# Number of samples
n_samp <- 3
# Default spawn index parameter values
data(pars) 
# Canada: coarse polygons for mapping
canada <- ne_countries(scale = "large", returnclass = "sf", country = "canada")
```

This example will focus on `r region`
from `r min(yr_range)` to `r max(yr_range)`.

## Spatial information

Load the spatial information for this region;
we will use the example spawn database that comes with
the package (`HerringSpawn.mdb`).
Note that this example database includes spawn survey data from 2010 to 2015
in the West Coast of Vancouver Island (WCVI) stock assessment region (SAR).

```{r areas}
# Database location
db_loc <- system.file("extdata", package = "SpawnIndex")
# List indicating the necessary tables: sections and locations
area_loc <- list(
  loc = db_loc,
  db = "HerringSpawn.mdb",
  fns = list(sections = "Sections", locations = "Location")
)
# Load area data
areas <- load_area_data(reg = region, where = area_loc)
# Determine spatial extent (for maps)
x_lims <- range(areas$Longitude)
y_lims <- range(areas$Latitude)
```

In this example database, the `r region` SAR has
`r n_distinct(areas$StatArea)` Statistical Areas,
`r n_distinct(areas$Section)` Sections, and
`r n_distinct(areas$LocationName)` Locations.
Here we show the first 10 Locations.

```{r areas_table, echo = FALSE, results = 'asis'}
# Table caption
areas_table_cap <- paste(
  "<i>Example spawn survey locations in the", region,
  "stock assessment region (SAR).</i>"
)
# Areas to include
areas %>%
  select(
    Region, StatArea, Section, LocationName, LocationCode, Longitude, Latitude
  ) %>%
  arrange(StatArea, Section, LocationName) %>%
  rename(
    `Statistical Area` = StatArea, `Location name` = LocationName,
    `Location code` = LocationCode
  ) %>%
  slice(1:10) %>%
  kable(
    caption = areas_table_cap, booktabs = TRUE, digits = c(0, 0, 0, 0, 0, 3, 3)
  )
```

## Spawn index calculations

Calculate the conversion factor to convert the number of eggs to
the spawn index (i.e., biomass in tonnes).

```{r theta}
# Egg to biomass conversion factor
theta <- eggs_to_sb()
```

Thus, we divide the number of eggs by $`r theta`$ to estimate
the biomass of fish that spawned.
This is the default conversion factor for the spawn index calculations below
(e.g., `calc_surf_index()`).
Then we calculate the spawn index for the three spawn survey types:
surface, Macrocystis, and understory.

### Surface spawn index

Calculate the surface spawn index.
Surface spawn index calculations require a table of median spawn width estimates
because width estimates from surfaces surveys are inaccurate.

```{r width}
# Median spawn width
width_loc <- list(
  loc = db_loc,
  db = "HerringSpawn.mdb",
  fns = list(
    region_std = "RegionStd", section_std = "SectionStd", pool_std = "PoolStd"
  )
)
width_bar <- load_width(where = width_loc, areas = areas)
```

```{r width_table, echo = FALSE, results = 'asis'}
# Table caption
width_table_cap <- paste(
  "<i>Median spawn width in metres (m) in the", region,
  "stock assessment region (SAR).</i>"
)
# Median spawn width: table
full_join(x = width_bar$region, y = width_bar$section, by = "Region") %>%
  full_join(y = width_bar$pool, by = c("Region", "Section")) %>%
  rename(
    `Region width` = WidthReg, `Section width` = WidthSec,
    `Pool width` = WidthPool
  ) %>%
  slice(1:10) %>%
  kable(
    caption = width_table_cap, booktabs = TRUE, digits = c(0, 1, 0, 1, 0, 1)
  )
```

Surface spawn index calculations also require a table to convert
intensity categories to the number of egg layers.

```{r surf}
# Spawn intensity categories and number of egg layers
data(intensity) 
# List indicating the necessary tables
surf_loc <- list(
  loc = db_loc,
  db = "HerringSpawn.mdb",
  fns = list(surface = "tSSSurface", all_spawn = "tSSAllspawn")
)
# Calculate the surface spawn index
surf_spawn <- calc_surf_index(
  where = surf_loc, areas = areas, widths = width_bar, years = yr_range
)
```

Show surface spawn on a map, aggregated by Location and year.

```{r surf_loc_yr, fig.cap = surf_loc_yr_cap}
# Figure caption
surf_loc_yr_cap <- paste(
  "<i>Surface spawn index in tonnes (t) by location and year from",
  min(yr_range), "to", max(yr_range), "in the", region,
  "stock assessment region (SAR).</i>"
)
# Make a spatial object for mapping
surf_loc_yr <- surf_spawn$si %>%
  group_by(Year, LocationCode) %>%
  summarise(Surface = round(sum(SurfSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = surf_loc_yr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Surface), size = 3) +
  scale_colour_viridis_c() +
  labs(colour = "Spawn\nindex (t)") +
  coord_sf(xlim = x_lims, ylim = y_lims) +
  facet_wrap(Year ~ ., ncol = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Calculate the total surface spawn by year.

```{r surf_yr, results = 'asis'}
# Table caption
surf_yr_cap <- paste(
  "<i>Surface spawn index in tonnes (t) by year from", min(yr_range), "to",
  max(yr_range), "in the", region, "stock assessment region (SAR).</i>"
)
# Aggregate the surface spawn index by year
surf_yr <- surf_spawn$si %>%
  group_by(Year) %>%
  summarise(Surface = sum(SurfSI, na.rm = TRUE)) %>%
  ungroup() %>%
  complete(Year = yr_range) %>%
  arrange(Year)
# Table
kable(x = surf_yr, digits = c(0, 1), booktabs = TRUE, caption = surf_yr_cap)
```

### Macrocystis spawn index

Calculate the Macrocystis spawn index.

```{r macro}
# List indicating the necessary tables
macro_loc <- list(
  loc = file.path(db_loc),
  db = "HerringSpawn.mdb",
  fns = list(
    all_spawn = "tSSAllspawn", plants = "tSSMacPlant", transects = "tSSMacTrans"
  )
)
# Calculate the Macrocystis spawn index
macro_spawn <- calc_macro_index(
  where = macro_loc, areas = areas, years = yr_range
)
```

Show Macrocystis spawn on a map, aggregated by Location and year.

```{r macro_loc_yr, fig.cap = macro_loc_yr_cap}
# Figure caption
macro_loc_yr_cap <- paste(
  "<i>Macrocystis spawn index in tonnes (t) by location and year from",
  min(yr_range), "to", max(yr_range), "in the", region,
  "stock assessment region (SAR).</i>"
)
# Make a spatial object for mapping
macro_loc_yr <- macro_spawn$si %>%
  group_by(Year, LocationCode) %>%
  summarise(Macrocystis = round(sum(MacroSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = macro_loc_yr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Macrocystis), size = 3) +
  scale_colour_viridis_c() +
  labs(colour = "Spawn\nindex (t)") +
  coord_sf(xlim = x_lims, ylim = y_lims) +
  facet_wrap(Year ~ ., ncol = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Calculate the total Macrocystis spawn by year.

```{r macro_yr, results = 'asis'}
# Table caption
macro_yr_cap <- paste(
  "<i>Macrocystis spawn index in tonnes (t) by year from", min(yr_range), "to",
  max(yr_range), "in the", region, "stock assessment region (SAR).</i>"
)
# Aggregate the Macrocystis spawn index by year
macro_yr <- macro_spawn$si %>%
  group_by(Year) %>%
  summarise(Macrocystis = sum(MacroSI, na.rm = TRUE)) %>%
  ungroup() %>%
  complete(Year = yr_range) %>%
  arrange(Year)
# Table
kable(x = macro_yr, digits = c(0, 1), booktabs = TRUE, caption = macro_yr_cap)
```

### Understory spawn index

Calculate the understory spawn index.
Understory spawn index calculations require tables to correct spawn width,
and to specify algae coefficients.

```{r under}
# Understory spawn width correction factors
data(under_width_facs)
# Algae types and coefficients
data(algae_coefs)
# List indicating the necessary tables
under_loc <- list(
  loc = file.path(db_loc),
  db = "HerringSpawn.mdb",
  fns = list(
    all_spawn = "tSSAllspawn", alg_trans = "tSSVegTrans",
    stations = "tSSStations", algae = "tSSVegetation"
  )
)
# Calculate the understory spawn index
under_spawn <- calc_under_index(
  where = under_loc, areas = areas, years = yr_range
)
```

Show understory spawn on a map, aggregated by Location and year.

```{r under_loc_yr, fig.cap = under_loc_yr_cap}
# Figure caption
under_loc_yr_cap <- paste(
  "<i>Understory spawn index in tonnes (t) by location and year from",
  min(yr_range), "to", max(yr_range), "in the", region,
  "stock assessment region (SAR).</i>"
)
# Make a spatial object for mapping
under_loc_yr <- under_spawn$si %>%
  group_by(Year, LocationCode) %>%
  summarise(Understory = round(sum(UnderSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = under_loc_yr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Understory), size = 3) +
  scale_colour_viridis_c() +
  labs(colour = "Spawn\nindex (t)") +
  coord_sf(xlim = x_lims, ylim = y_lims) +
  facet_wrap(Year ~ ., ncol = 2) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Calculate the total understory spawn by year.

```{r under_yr, results = 'asis'}
# Table caption
under_yr_cap <- paste(
  "<i>Understory spawn index in tonnes (t) by year from", min(yr_range), "to",
  max(yr_range), "in the", region, "stock assessment region (SAR).</i>"
)
# Aggregate the understory spawn index by year
under_yr <- under_spawn$si %>%
  group_by(Year) %>%
  summarise(Understory = sum(UnderSI, na.rm = TRUE)) %>%
  ungroup() %>%
  complete(Year = yr_range) %>%
  arrange(Year)
# Table
kable(x = under_yr, digits = c(0, 1), booktabs = TRUE, caption = under_yr_cap)
```

### Total spawn index

Calculate the total spawn index by year as the sum of spawn index
from the three survey types:
surface, Macrocystis, and understory.

```{r total_yr, results = 'asis'}
# Table caption
total_yr_cap <- paste(
  "<i>Total spawn index in tonnes (t) by year from", min(yr_range), "to",
  max(yr_range), "in the", region, "stock assessment region (SAR).</i>"
)
# Total spawn index
total_yr <- full_join(x = surf_yr, y = macro_yr, by = "Year") %>%
  full_join(y = under_yr, by = "Year") %>%
  group_by(Year) %>%
  mutate(Total = sum(c(Surface, Macrocystis, Understory), na.rm = TRUE)) %>%
  ungroup() %>%
  complete(Year = yr_range) %>%
  arrange(Year)
# Table
kable(
  x = total_yr, digits = c(0, 1, 1, 1, 1), booktabs = TRUE,
  caption = total_yr_cap
) %>%
  add_header_above(header = c("", "Survey" = 3, ""))
```

# Surface spawn index calculations using custom data

Calculate the Pacific Herring spawn index using custom data,
such as would be collected during a surface spawn survey.
<!-- First we provide an example where surveyors did surface surveys, -->
<!-- then we provide examples where surveyors did underwater dive surveys. -->
First, generate some values for the number of egg layers in each sample,
where there are `r n_samp` samples in each of the `r length(yr_range)` spawns.

```{r surface_raw}
# Generate some data: samples within spawns
surf_raw <- tibble(
  Spawn = rep(1:length(yr_range), each = n_samp),
  Sample = rep(1:n_samp, times = length(yr_range)),
  Layers = sample(1:9, size = length(yr_range) * n_samp, replace = TRUE)
)
# Table caption
surf_raw_cap <- paste(
  "<i>Surface spawn survey observations: number of egg layers in each",
  "sample.</i>"
)
# Table
kable(
  x = surf_raw, digits = c(0, 0, 0), booktabs = TRUE, caption = surf_raw_cap
)
```

Then we summarise each spawn by the average number of egg layers,
as well as the length and width of the spawn in metres.

```{r surace_dat}
# Summarise the data
surf_dat <- surf_raw %>%
  group_by(Spawn) %>%
  summarise(Layers = mean(Layers)) %>%
  ungroup() %>%
  mutate(
    Length = sample(x = 1000:5000, size = length(yr_range)),
    Width = sample(x = 20:100, size = length(yr_range))
  )
# Table caption
surf_dat_cap <- paste(
  "<i>Surface spawn survey observations: average number of egg layers, as well",
  "as length and width of the spawn in metres (m).</i>"
)
# Table
kable(
  x = surf_dat, digits = c(0, 1, 0, 0), booktabs = TRUE, caption = surf_dat_cap
)
```

Finally, we calculate egg density using the default parameters, and
calculate the spawn index in tonnes for each spawn.

```{r surface_index}
# Calculate the index
surf_ind <- surf_dat %>%
  mutate(
    Density = dens_surf(egg_layers = Layers),
    Index = Density * Length * Width * 1000 / theta
  )
# Table caption
surf_ind_cap <- paste(
  "<i>Surface spawn survey observations and spawn index: average number of egg",
  "layers, length and width of the spawn in metres (m), egg density in",
  "thousands of eggs per square metre (10^3 * eggs / m^2), and surface spawn",
  "index in tonnes (t).</i>"
)
# Table
kable(
  x = surf_ind, digits = c(0, 1, 0, 0, 0, 1), booktabs = TRUE,
  caption = surf_ind_cap
)
```

<!-- ## Macrocystis spawn index -->

<!-- Similar to the previous example, -->
<!-- we assume that there were `r length(yr_range)` spawns -->
<!-- which are summarised by -->
<!-- average number of egg layers, -->
<!-- average plant height in metres, -->
<!-- average number of stalks per plant, as well as -->
<!-- the length and width of the spawn in metres. -->
<!-- We also calculate the number of eggs per plant using the default parameters, -->
<!-- and then calculate the spawn index in tonnes for each spawn. -->

```{r macrocystis, include = FALSE, eval = FALSE}
# Generate some data
macro_dat <- tibble(
  Number = 1:length(yr_range),
  Layers = sample(x = 2:8, size = length(yr_range), replace = TRUE),
  Height = sample(x = 2:10, size = length(yr_range), replace = TRUE),
  StalksPerPlant = sample(x = 2:5, size = length(yr_range), replace = TRUE),
  Plants = sample(x = 5:20, size = length(yr_range))
)
# Calculate the index
macro_ind <- macro_dat %>%
  mutate(
    EggsPerPlant = eggs_macro(
      egg_layers = Layers, height = Height, stalks_per_plant = StalksPerPlant
    ),
    Eggs = EggsPerPlant * Plants,
    Index = Eggs * 1000 / theta
  )
# Table caption
macro_ind_cap <- paste(
  "<i>Macrocystis spawn survey observations and spawn index: average number of",
  "egg layers, average plant height in metres (m), average number of stalks",
  "per plant, number of eggs per plant in thousands (10^3 * eggs / plant),",
  "number of eggs in thousands (10^3 * eggs), and macrocystis spawn index in",
  "tonnes (t).</i>"
)
# Table
kable(
  x = macro_ind, digits = c(0, 0, 0, 0, 0, 1, 1, 1), booktabs = TRUE,
  caption = macro_ind_cap
)
```

Note that this example is for surface spawn surveys,
but Macrocystis and understory spawn surveys would follow similar processes.

# Spawn index calculations for spawn-on-kelp (SOK) operations

Calculate the biomass of spawning Pacific Herring from
spawn-on-kelp (SOK) operations.
Note that these are fictional values.

```{r sok_yr, results='asis'}
# Table caption
sok_yr_cap <- paste(
  "<i>SOK harvest in kilograms (kg) by year and associated biomass in tonnes",
  "(t) from", min(yr_range), "to", max(yr_range), "in the", region,
  "stock assessment region (SAR).</i>"
)
# Set a random seed
set.seed(12)
# SOK harvest (kg) by year -- make up some values
kg_sok <- tibble(
  Year = yr_range, Harvest = sample(x = 500:10000, size = length(yr_range))
) %>%
  mutate(Biomass = calc_sok_index(Harvest))
# Table
kable(x = kg_sok, digits = c(0, 0, 1), booktabs = TRUE, caption = sok_yr_cap)
```

# References
