---
title: "Pacific Herring spawn index: An example"
author: "Matthew H. Grinnell"
date: "`r Sys.Date()`"
output:
      rmarkdown::html_vignette:
        fig_caption: yes
        toc: true
bibliography: "../inst/REFERENCES.bib"
link-citations: true
vignette: >
  %\VignetteIndexEntry{Pacific Herring spawn index: An example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)
options(knitr.kable.NA = "")
```

This vignette provides an example on how to calculate the spawn index for Pacific Herring (*Clupea pallasii*) in British Columbia, Canada.
See the technical report for background information [@GrinnellEtalYYYY].
Source code is on the
[GitHub repository](https://github.com/grinnellm/SpawnIndex).
Load the `SpawnIndex` package and a few other packages for analyses, tables, and figures.

```{r libraries, message = FALSE, warning = FALSE}
library(SpawnIndex)
library(dplyr)
library(ggplot2)
library(tidyr)
library(sf)
library(kableExtra)
library(rnaturalearth)
library(rnaturalearthhires)
```

# Set up

Specify the region and years of interest.
Also, load the parameter values required for spawn index calculations.
Finally, load some land polygons for maps

```{r pars}
# Parameters
region <- "WCVI" # Region: HG, PRD, CC, SoG, WCVI, A27, or A2W
yrRange <- 2010:2015 # Years (>= 1951)
data(pars) # Parameters for spawn index calculations
# Canada: coarse polygons for mapping
canada <- ne_countries(scale = "large", returnclass = "sf", country = "canada")
```

This example will focus on `r region` from `r min(yrRange)` to `r max(yrRange)`.

# Spatial information

Load the spatial information for this region;
we will use the example spawn database that comes with the package (`HerringSpawn.mdb`).
Note that this example database only includes WCVI from 2010 to 2015.

```{r areas}
# Database location
dbLoc <- system.file("extdata", package = "SpawnIndex")
# List indicating the necessary tables: sections and locations
areaLoc <- list(
  loc = dbLoc, db = "HerringSpawn.mdb",
  fns = list(sections = "Sections", locations = "Location")
)
# Load area data
areas <- LoadAreaData(reg = region, where = areaLoc)
# Determine spatial extent (for maps)
xLims <- range(areas$Longitude)
yLims <- range(areas$Latitude)
```

In this example database, `r region` has
`r n_distinct(areas$StatArea)` Statistical Areas,
`r n_distinct(areas$Section)` Sections, and
`r n_distinct(areas$LocationName)` Locations.
Here we show the first 10 Locations.

```{r areasTable, echo = FALSE, results = 'asis'}
areas %>%
  select(Region, StatArea, Section, LocationName, Longitude, Latitude) %>%
  arrange(StatArea, Section, LocationName) %>%
  slice(1:10) %>%
  kable(
    caption = paste("Example spawn survey Locations in the", region, "region."),
    booktabs = TRUE
  )
```

# Spawn index calculations

Calculate the conversion factor to convert the number of eggs to the spawn index (i.e., biomass in tonnes).

```{r theta}
# Egg to biomass conversion factor
theta <- CalcEggConversion()
```

Thus, we divide the number of eggs by $`r theta`$ to estimate the biomass of fish that spawned.
This is the default conversion factor for the spawn index calculations below (e.g., `CalcSurfSpawn()`).
Then we calculate the spawn index for the three spawn survey types:
surface, Macrocystis, and understory.

## Surface spawn index

Calculate the surface spawn index.
Surface spawn index calculations require a table of median spawn width estimates because width estimates from surfaces surveys are inaccurate.

```{r width}
widthLoc <- list(
  loc = dbLoc, db = "HerringSpawn.mdb",
  fns = list(
    regionStd = "RegionStd", sectionStd = "SectionStd", poolStd = "PoolStd"
  )
)
barWidth <- GetWidth(where = widthLoc, a = areas)
```

```{r widthTable, echo = FALSE, results = 'asis'}
full_join(
  x = barWidth$region,
  y = barWidth$section, by = "Region"
) %>%
  full_join(y = barWidth$pool, by = c("Region", "Section")) %>%
  rename(
    `Region width` = WidthReg, `Section width` = WidthSec,
    `Pool width` = WidthPool
  ) %>%
  slice(1:10) %>%
  kable(
    caption = paste(
      "Median spawn widths in metres (m) in the", region, "region."
    ),
    booktabs = TRUE
  )
```

Surface spawn index calculations also require a table to convert intensity categories to the number of egg layers.

```{r surf}
# Data
data(intensity) # Spawn intensity categories and number of egg layers
# List indicating the necessary tables
surfLoc <- list(
  loc = dbLoc, db = "HerringSpawn.mdb",
  fns = list(surface = "tSSSurface", allSpawn = "tSSAllspawn")
)
# Calculate the surface spawn index
surfSpawn <- CalcSurfSpawn(
  where = surfLoc, a = areas, widths = barWidth, yrs = yrRange
)
```

Show surface spawn on a map, aggregated by Location and year.

```{r surfLocYr, fig.cap = "Surface spawn index in tonnes (t) by location and year."}
# Make a spatial object for mapping
surfLocYr <- surfSpawn$SI %>%
  group_by(Year, LocationCode) %>%
  summarise(Surface = round(sum(SurfSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = surfLocYr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Surface), size = 3) +
  scale_colour_viridis_c() +
  labs(colour = "Spawn\nindex (t)") +
  coord_sf(xlim = xLims, ylim = yLims) +
  facet_wrap(Year ~ ., ncol = 2) +
  theme_bw()
```

Calculate the total surface spawn by year.

```{r surfYr, results = 'asis'}
# Aggregate the surface spawn index by year
surfYr <- surfSpawn$SI %>%
  group_by(Year) %>%
  summarise(Surface = sum(SurfSI, na.rm = TRUE)) %>%
  ungroup() %>%
  complete(Year = yrRange) %>%
  arrange(Year)
# Table
kable(
  x = surfYr, digits = 1, booktabs = TRUE,
  caption = "Surface spawn index in tonnes (t) by year."
)
```

## Macrocystis spawn index

Calculate the Macrocystis spawn index.

```{r macro}
# List indicating the necessary tables
macroLoc <- list(
  loc = file.path(dbLoc), db = "HerringSpawn.mdb",
  fns = list(
    allSpawn = "tSSAllspawn", plants = "tSSMacPlant", transects = "tSSMacTrans"
  )
)
# Calculate the Macrocystis spawn index
macroSpawn <- CalcMacroSpawn(where = macroLoc, a = areas, yrs = yrRange)
```

Show Macrocystis spawn on a map, aggregated by Location and year.

```{r macroLocYr, fig.cap = "Macrocystis spawn index in tonnes (t) by location and year."}
# Make a spatial object for mapping
macroLocYr <- macroSpawn$SI %>%
  group_by(Year, LocationCode) %>%
  summarise(Macrocystis = round(sum(MacroSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = macroLocYr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Macrocystis), size = 3) +
  scale_colour_viridis_c() +
  labs(colour = "Spawn\nindex (t)") +
  coord_sf(xlim = xLims, ylim = yLims) +
  facet_wrap(Year ~ ., ncol = 2) +
  theme_bw()
```

Calculate the total Macrocystis spawn by year.

```{r macroYr, results = 'asis'}
# Aggregate the Macrocystis spawn index by year
macroYr <- macroSpawn$SI %>%
  group_by(Year) %>%
  summarise(Macrocystis = sum(MacroSI, na.rm = TRUE)) %>%
  ungroup() %>%
  complete(Year = yrRange) %>%
  arrange(Year)
# Table
kable(
  x = macroYr, digits = 1, booktabs = TRUE,
  caption = "Macrocystis spawn index in tonnes (t) by year."
)
```

## Understory spawn index

Calculate the understory spawn index.
Understory spawn index calculations require tables to correct spawn widths,
and to specify algae coefficients.

```{r under}
# Data
data(underWidthFac) # Understory spawn width correction factors
data(algaeCoefs) # Algae types and coefficients
# List indicating the necessary tables
underLoc <- list(
  loc = file.path(dbLoc), db = "HerringSpawn.mdb",
  fns = list(
    allSpawn = "tSSAllspawn", algTrans = "tSSVegTrans",
    stations = "tSSStations", algae = "tSSVegetation"
  )
)
# Calculate the understory spawn index
underSpawn <- CalcUnderSpawn(where = underLoc, a = areas, yrs = yrRange)
```

Show understory spawn on a map, aggregated by Location and year.

```{r underLocYr, fig.cap = "Understory spawn index in tonnes (t) by location and year."}
# Make a spatial object for mapping
underLocYr <- underSpawn$SI %>%
  group_by(Year, LocationCode) %>%
  summarise(Understory = round(sum(UnderSI, na.rm = TRUE))) %>%
  ungroup() %>%
  left_join(
    y = select(.data = areas, LocationCode, Longitude, Latitude),
    by = "LocationCode"
  ) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)
# Map
ggplot(data = underLocYr) +
  geom_sf(data = canada) +
  geom_sf(mapping = aes(colour = Understory), size = 3) +
  scale_colour_viridis_c() +
  labs(colour = "Spawn\nindex (t)") +
  coord_sf(xlim = xLims, ylim = yLims) +
  facet_wrap(Year ~ ., ncol = 2) +
  theme_bw()
```

Calculate the total understory spawn by year.

```{r underYr, results = 'asis'}
# Aggregate the understory spawn index by year
underYr <- underSpawn$SI %>%
  group_by(Year) %>%
  summarise(Understory = sum(UnderSI, na.rm = TRUE)) %>%
  ungroup() %>%
  complete(Year = yrRange) %>%
  arrange(Year)
# Table
kable(
  x = underYr, digits = 1, booktabs = TRUE,
  caption = "Understory spawn index in tonnes (t) by year."
)
```

## Total spawn index

Calculate the total spawn index by year as the sum of the spawn index in the three survey types:
surface, Macrocystis, and understory.

```{r totalYr, results = 'asis'}
# Total spawn index
totalYr <- full_join(x = surfYr, y = macroYr, by = "Year") %>%
  full_join(y = underYr, by = "Year") %>%
  group_by(Year) %>%
  mutate(Total = sum(c(Surface, Macrocystis, Understory), na.rm = TRUE)) %>%
  ungroup() %>%
  complete(Year = yrRange) %>%
  arrange(Year)
# Table
kable(
  x = totalYr, digits = 1, booktabs = TRUE,
  caption = "Total spawn index in tonnes (t) by year."
) %>%
  add_header_above(header = c("", "Survey" = 3, ""))
```

# Spawn on kelp calculations

Calculate the biomass of spawning Pacific Herring from spawn on kelp (SOK) operations.
Note that these are fictional values.

```{r sokYr, results='asis'}
# SOK harvest (kg) by year -- make up some values
kgSOK <- tibble(
  Year = yrRange,
  Harvest = sample(x = 500:10000, size = length(yrRange))
) %>%
  mutate(Biomass = CalcBiomassSOK(Harvest))
# Table
kable(
  x = kgSOK, digits = 1, booktabs = TRUE,
  caption = paste(
    "SOK harvest in kilograms (kg) by year and",
    "associated biomass in tonnes (t)."
  )
)
```


# References
